---
title: "lifequal: Calculating Life-table Lifespan Equality"
author: "Jonas Schöley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lifequal: Calculating Lifespan Equality from Life-tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

How to install?
---------------

You can install `lifequal` by running:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("jschoeley/lifequal")
```

What does it do?
----------------

`lifequal` lets you calculate three measures of lifespan equality from a life-table:

  1) `ExDagger(x, ex, wx, ax)`
  Life expectancy lost by those who die in age interval [x, x+w)
  2) `EDagger(dx, exdagger, radix)`
  Total life expectancy lost due to death
  3) `KeyfzEntro(edagger, e0)`
  Keyfitz's entropy

They are defined as follows,

--------------------------------------------------------------------------------
Measure                                    Definition
------------------------------------------ -------------------------------------
Start of age interval                      $x$

Width of age interval starting at $x$      $w_x$

Start of last age interval                 $\omega$

Average time spent in age interval         $a_x$
$[x, x+wx)$ when dying in that interval

Deaths in age interval $[x, x+w_x)$        $d_x$

Life-expectancy at age $x$                 $e_x$

Life expectancy lost due to death in age   $e_x^\dagger = \frac {a_x} {w_x}e_{x+w_x} + (1-\frac {a_x} {w_x}) e_x$
interval $[x, x+w_x)$

Total life expectancy lost due to death    $e^\dagger = \sum_{x=0}^{\omega} d_x e_x^\dagger$

Keyfitz's Entropy                          $\frac{e^\dagger} {e_0}$
--------------------------------------------------------------------------------

Life expectancy versus lifespan equality for 1x1 Swedish life-tables
--------------------------------------------------------------------

```{r, message=FALSE, warning=FALSE}
library(lifequal)
library(dplyr)
library(ggplot2)
```

The analysis starts with a demographic life-table. We want 1) age groups ordered from low to high, 2) no gaps between subsequent age groups. Something like this:

```{r}
# Swedish 1x1 period life-tables by period and sex
sweden1x1
```

First, we use `ExDagger()` on each single life-table (separate by period and sex) to calculate the life expectancy lost in each age. We then summarise each life-table into a set of 3 numbers: Life expectancy at birth, total life years lost due to death (`EDagger()`) and lifespan equality (`KeyfzEntro()`). Note that we transform Keyfitz's Entropy by taking the negative log.

```{r}
sweden1x1 %>%
  # ...for each single life-table...
  group_by(period, sex) %>%
  #...we calculate the life years lost in age x...
  mutate(exdagger = ExDagger(x, ex)) %>%
  # ...and then summarise each life-table into a set of 3 numbers:
  # e0:         Life-expectancy at birth
  # edagger:    Total life years lost due to death
  # keyfzentro: Lifespan equality
  summarise(
    e0         = ex[x == 0],
    edagger    = EDagger(dx, exdagger, radix = 100000),
    keyfzentro = -log(KeyfzEntro(edagger, e0))
  ) -> sweden1x1summary
```

The summarised life-tables look like this:

```{r}
sweden1x1summary
```

For each life-table we plot the life expectancy at birth versus the lifespan equality.

```{r, fig.height=5, fig.width=5}
plot_lifequal <-
  ggplot(sweden1x1summary, aes(x = e0, y = keyfzentro, color = sex)) +
  geom_point(size = 0.6, alpha = 0.5) +
  theme(aspect.ratio = 1)
plot_lifequal
```

Life expectancy versus lifespan equality for 5x5 Swedish life-tables
--------------------------------------------------------------------

The same exercise as before, only now we deal with life-tables aggregated over multiple year period and age intervals.

```{r}
# Swedish 5x5 period life-tables by period and sex
sweden5x5
```

We deal with this aggregation by informing `ExDagger()` about $w_x$, the width of each age interval. This information is included in the column `wx` of our life-table.

```{r}
sweden5x5 %>%
  # for practicality we assume a width of 2
  # years for the last open age group 110+
  mutate(wx = ifelse(is.na(wx), 2, wx)) %>%
  # ...for each single life-table...
  group_by(period, sex) %>%
  #...we calculate the life years lost in age interval [x, x+wx)...
  mutate(exdagger = ExDagger(x, ex, wx, ax)) %>%
  # ...and then summarise each life-table into a set of 3 numbers:
  # e0:         Life-expectancy at birth
  # edagger:    Total life years lost due to death
  # keyfzentro: Lifespan equality
  summarise(
    e0         = ex[x == 0],
    edagger    = EDagger(dx, exdagger, radix = 100000),
    keyfzentro = -log(KeyfzEntro(edagger, e0))
  ) -> sweden5x5summary
```

The summarised life-tables look like this:

```{r}
sweden5x5summary
```

The aggregated life-tables follow the same trend as the single year life-tables.

```{r, fig.height=5, fig.width=5}
plot_lifequal +
  geom_point(shape = 22, fill = "white",
             data = sweden5x5summary)
```
